---
title: "Chapter_9"
author: "Manuel Kunz"
date: "2023-10-04"
output: html_document
---

```{r, setup, echo=FALSE, include=FALSE}
library(dplyr)
library(readr)
library(lubridate)
library(ggplot2)
library(here)
library(caret)
library(tidyr)
```

## Data and the modelling challenge
In this chapter, we formulate a model for predicting GPP from a set of covariates (other variables that vary in parallel, here the meteorological variables). This is to say that GPP_NT_VUT_REF is the target variable, and other variables that are available in our dataset are the predictors.

Let’s read the data, select suitable variables, interpret missing value codes, and select only good-quality data (where at least 80% of the underlying half-hourly data was good quality measured data, and not gap-filled).

### Data wrangling code

```{r,data wrangling,message=FALSE}
daily_fluxes <- readr::read_csv(paste0(here::here(),"./data/FLX_CH-Dav_FLUXNET2015_FULLSET_DD_1997-2014_1-3.csv")) |>

  # select only the variables we are interested in
  dplyr::select(TIMESTAMP,
                GPP_NT_VUT_REF,    # the target
                ends_with("_QC"),  # quality control info
                ends_with("_F"),   # includes all all meteorological covariates
                -contains("JSB")   # weird useless variable
                ) |>

  # convert to a nice date object
  dplyr::mutate(TIMESTAMP = lubridate::ymd(TIMESTAMP)) |>

  # set all -9999 to NA
  mutate(across(where(is.numeric), ~na_if(., -9999))) |> 
  
  # retain only data based on >=80% good-quality measurements
  # overwrite bad data with NA (not dropping rows)
  dplyr::mutate(GPP_NT_VUT_REF = ifelse(NEE_VUT_REF_QC < 0.8, NA, GPP_NT_VUT_REF),
                TA_F           = ifelse(TA_F_QC        < 0.8, NA, TA_F),
                SW_IN_F        = ifelse(SW_IN_F_QC     < 0.8, NA, SW_IN_F),
                LW_IN_F        = ifelse(LW_IN_F_QC     < 0.8, NA, LW_IN_F),
                VPD_F          = ifelse(VPD_F_QC       < 0.8, NA, VPD_F),
                PA_F           = ifelse(PA_F_QC        < 0.8, NA, PA_F),
                P_F            = ifelse(P_F_QC         < 0.8, NA, P_F),
                WS_F           = ifelse(WS_F_QC        < 0.8, NA, WS_F)) |> 

  # drop QC variables (no longer needed)
  dplyr::select(-ends_with("_QC"))

```

## Model formulation
### Formula notation
In R, it is common to use the formula notation to specify the target and predictor variables. You have encountered formulas before, e.g., for a linear regression using the lm() function. To specify a linear regression model for GPP_NT_VUT_REF with three predictors SW_IN_F, VPD_F, and TA_F, to be fitted to data daily_fluxes, we write:
```{r, linear regression example, message = FALSE}
lm(GPP_NT_VUT_REF ~ SW_IN_F + VPD_F + TA_F, data = daily_fluxes)
```
### Using the caret package to call the base-R regression function
```{r, caret regression, message=FALSE}
caret::train(
  form = GPP_NT_VUT_REF ~ SW_IN_F + VPD_F + TA_F,
  data = daily_fluxes |> drop_na(), # drop missing values
  trControl = caret::trainControl(method = "none"), # no resampling
  method = "knn"
)
```
```{r, message = FALSE}
set.seed(123)  # for reproducibility
split <- rsample::initial_split(daily_fluxes, prop = 0.7, strata = "VPD_F")
daily_fluxes_train <- rsample::training(split)
daily_fluxes_test <- rsample::testing(split)

plot_data <- daily_fluxes_train |> 
  dplyr::mutate(split = "train") |> 
  dplyr::bind_rows(daily_fluxes_test |> 
  dplyr::mutate(split = "test")) |> 
  tidyr::pivot_longer(cols = 2:9, names_to = "variable", values_to = "value")

plot_data |> 
  ggplot(aes(x = value, y = ..density.., color = split)) +
  geom_density() +
  facet_wrap(~variable, scales = "free")

```
```{r}
pp <- recipes::recipe(GPP_NT_VUT_REF ~ SW_IN_F + VPD_F + TA_F, data = daily_fluxes_train) |> 
  recipes::step_center(recipes::all_numeric(), -recipes::all_outcomes()) |>
  recipes::step_scale(recipes::all_numeric(), -recipes::all_outcomes())

caret::train(
  pp, 
  data = daily_fluxes_train, 
  method = "knn",
  trControl = caret::trainControl(method = "none")
)
```

### Standardisation

```{r}
daily_fluxes |> 
  summarise(across(where(is.numeric), ~quantile(.x, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE))) |> 
  t() |> 
  as_tibble(rownames = "variable") |> 
  setNames(c("variable", "min", "q25", "q50", "q75", "max"))
```
####
```{r, data preparation, message=FALSE}
pp_prep <- recipes::prep(pp, training = daily_fluxes_train) 
```

#### transform data
```{r, transform data, message=FALSE}
daily_fluxes_juiced <- recipes::juice(pp_prep)
```

#### bake data
```{r, bake data, message = FALSE}
daily_fluxes_baked <- recipes::bake(pp_prep, new_data = daily_fluxes_train)

# confirm that juice and bake return identical objects when given the same data
all_equal(daily_fluxes_juiced, daily_fluxes_baked)
```
#### Plot original and standardasied data
```{r, standardised data plots, message = FALSE}
# prepare data for plotting
plot_data_original <- daily_fluxes_train |> 
  dplyr::select(one_of(c("SW_IN_F", "VPD_F", "TA_F"))) |> 
  tidyr::pivot_longer(cols = c(SW_IN_F, VPD_F, TA_F), names_to = "var", values_to = "val")

plot_data_juiced <- daily_fluxes_juiced |> 
  dplyr::select(one_of(c("SW_IN_F", "VPD_F", "TA_F"))) |> 
  tidyr::pivot_longer(cols = c(SW_IN_F, VPD_F, TA_F), names_to = "var", values_to = "val")

# plot density
plot_1 <- ggplot(data = plot_data_original, aes(val, ..density..)) +
  geom_density() +
  facet_wrap(~var)

# plot density by var
plot_2 <- ggplot(data = plot_data_juiced, aes(val, ..density..)) +
  geom_density() +
  facet_wrap(~var)

# combine both plots
cowplot::plot_grid(plot_1, plot_2, nrow = 2)
```
### Data gaps
#### visualise missing data
````{r, visualise missing data, message = FALSE}
visdat::vis_miss(
  daily_fluxes,
  cluster = FALSE, 
  warn_large_data = FALSE
  )
```

#### zero variance predictor
```{r, zero variance predicator, message = FALSE}
caret::nearZeroVar(daily_fluxes, saveMetrics = TRUE)
```
#### target engineering
```{r, target engineering, message = FALSE}
plot_1 <- ggplot(data = daily_fluxes, aes(x = WS_F, y = ..density..)) +
  geom_histogram() +
  labs(title = "Original")

plot_2 <- ggplot(data = daily_fluxes, aes(x = log(WS_F), y = ..density..)) +
  geom_histogram() +
  labs(title = "Log-transformed")

cowplot::plot_grid(plot_1, plot_2)
```
```{r}
recipes::recipe(WS_F ~ ., data = daily_fluxes) |>   # it's of course non-sense to model wind speed like this
  recipes::step_log(all_outcomes())
```
#### Box-Cox transformation
```{r box-cox-transformation, message=FALSE}
pp <- recipes::recipe(WS_F ~ ., data = daily_fluxes_train) |>
  recipes::step_BoxCox(recipes::all_outcomes())

prep_pp <- recipes::prep(pp, training = daily_fluxes_train |> drop_na())
daily_fluxes_baked <- recipes::bake(prep_pp, new_data = daily_fluxes_test |> drop_na())
daily_fluxes_baked |>
  ggplot(aes(x = WS_F, y = ..density..)) +
  geom_histogram() +
  labs(title = "Box-Cox-transformed")
```
### Summary
#### Data cleaning
```{r}
# Data cleaning: looks ok, no obviously bad data
# no long tail, therefore no further target engineering
daily_fluxes |> 
  ggplot(aes(x = GPP_NT_VUT_REF, y = ..count..)) + 
  geom_histogram()
```

#### Model setup
```{r, model setup}
# Data splitting
set.seed(1982)  # for reproducibility
split <- rsample::initial_split(daily_fluxes, prop = 0.7, strata = "VPD_F")
daily_fluxes_train <- rsample::training(split)
daily_fluxes_test <- rsample::testing(split)

# Model and pre-processing formulation, use all variables but LW_IN_F
pp <- recipes::recipe(GPP_NT_VUT_REF ~ SW_IN_F + VPD_F + TA_F, 
                      data = daily_fluxes_train |> drop_na()) |> 
  recipes::step_BoxCox(recipes::all_predictors()) |> 
  recipes::step_center(recipes::all_numeric(), -recipes::all_outcomes()) |>
  recipes::step_scale(recipes::all_numeric(), -recipes::all_outcomes())

# Fit linear regression model
mod_lm <- caret::train(
  pp, 
  data = daily_fluxes_train |> drop_na(), 
  method = "lm",
  trControl = caret::trainControl(method = "none"),
  metric = "RMSE"
)

# Fit KNN model
mod_knn <- caret::train(
  pp, 
  data = daily_fluxes_train |> drop_na(), 
  method = "knn",
  trControl = caret::trainControl(method = "none"),
  tuneGrid = data.frame(k = 8),
  metric = "RMSE"
)
```
#### model evaluation
```{r, model evaluation}
source(paste0(here::here(),"./R/eval_model.R"))

# linear regression model
eval_model(mod = mod_lm, df_train = daily_fluxes_train, df_test = daily_fluxes_test)
```
#### KNN model evaluation
```{r, KNN model}
eval_model(mod = mod_knn, df_train = daily_fluxes_train, df_test = daily_fluxes_test)
```
